Density-dependent selection and the limits of relative fitness

### 2017-12-06 manuscript.tex
l

\title{Density-dependent selection and the limits of relative fitness}
%\author{Jason Bertram $^{1,\ast}$ \\ 
%Joanna Masel $^{1}$}

@@

\textit{Keywords}: $r$/$K$ selection, absolute fitness, eco-evo, competition-colonization trade-off, fluctuating selection, storage effect.

@@

new model of density-dependent selection by generalizing the fixed-density classic lottery model of territorial acquisition to accommodate arbitrary population densities.

We show that, with density dependence, co-existence is possible in the lottery model in a stable environment.

Inspired by natural \textit{Drosophila} populations, we consider co-existence under strong, seasonally-fluctuating selection coupled to large cycles in population density, and show that co-existence (stable polymorphism) is promoted via a combination of the classic storage effect and density-regulated population growth.

@@

In uncrowded populations, relative fitness simply represents differences in intrinsic growth rate. In continuous time, we have the canonical selection equation $\frac{d p_i}{dt}=(r_i-\overline{r})p_i$, where $r_i$ is the intrinsic exponential growth rate of type $i$, $p_i$ is its frequency, and $\overline{r}=\sum_i r_i p_i$ is the population mean $r$ \citep[pp. 26]{crow_1970}. If there are two types present, a wildtype and a mutant for instance, then Eq.~\eqref{eq:canonical} can be written as 
\begin{equation}
\frac{d p_i}{dt}=s p_i(1-p_i), \label{eq:canonical}
\end{equation}
where the constant selection coefficient $s$ is the difference in $r$ between types. The corresponding adaptive sweeps follow a logistic curve.

The difficulty with Eq.~\eqref{eq:canonical} arises in crowded populations. It is then commonly assumed that total population density $N$ has reached its equilibrium value, which is assumed to be a fixed constant. This is exemplified by the Wright-Fisher model. Yet MacArthur famously showed that when population growth is density-regulated, selection in crowded populations is intimately connected to the ability to keep growing at higher densities than other types can tolerate \citep{macarthur_1967}. The classic example is the logistic model, where the type with the greatest saturation population density ``$K$'' excludes the others (Fig.~\ref{fig:Ksel}a). Similarly, the ``$R^*$ rule'', a central tenet of resource competition theory, states that when growth is limited by a single homogeneous consumable resource, the type able to deplete the resource to the lowest equilibrium density $R^*$ excludes the others \citep{grover_1997}. Differences in $R^*$ will often entail differences in saturation density. The Lotka-Volterra competition model also couples selection in crowded populations to changes in $N$ except in a few special cases (we return to this in section ``$K$-selection, $c$-selection and relative fitness''). In these examples, both $N$ and $s$ change during, and as a result of, adaptive sweeps. It would therefore seem that the ubiquitous constant-$N$, relative fitness description of selection is incompatible with a huge class of population ecological processes occurring in nature and experiments (Fig.~\ref{fig:Ksel}b).

@@

The relative fitness description has been justified in broadly two different ways for crowded populations (we do not address Wagner's [\citeyear{wagner_2010}] measure theorical justification, which explicitly seeks to be independent of population biology and thus falls outside of our scope). First, if we only care about type frequencies, we can sometimes simply ignore $N$ \cite[pp. 468]{barton_2007}, because Eq.~\eqref{eq:canonical} is unaffected by changes in $N$ if the mechanism regulating density applies equally to all types \citep{prout_1980}. While this does allow us to relax the restrictive assumption of constant $N$, it does not address the assumption of constant $s$. In the examples from the previous paragraph, different types are affected by the regulation of density in different ways; indeed, the type-specific responses to density  are at the center of MacArthur's argument and the literature on density-dependent selection that grew out of it.

Second, constant $N$ and $s$ can both be seen as a short-term linear approximation \cite[pp. 277]{ewens_2004}. That is, within a sufficiently short time frame, $N$ and $s$ can be treated as constant. Provided that selection is sufficiently weak and stable over time, this ``short-term''  assumption is not a major restriction. Yet it is increasingly recognized that selection is not always weak, that it can fluctuate considerably over time, and that $N$ can vary by orders of magnitude over a few generations as a routine feature of a population's ecology \citep{messer_2016}. These are not rare exceptions, but occur widely in nature and the lab, including in wild \textit{Drosophila} \citep{bergland_14}. Nevertheless, relative fitness models like Wright-Fisher are the foundation for much of the population genetic literature, and are still widely used without considering the ``short-term'' restriction or the gulf with population ecology \citep{mallet_2012}. Thus, it is important to understand the population ecological basis of relative fitness models, both to gain insight into their domain of applicability, and as part of the broader challenge of synthesizing ecology and evolution.

Here we analyze the population ecology of relative fitness using a novel model of density-dependent population growth based on territorial contests. Rather than starting from the standard models of population growth mentioned above, our point of departure  is the classic lottery model \citep{sale_77,chesson_1981}. Like the Wright-Fisher model, the classic lottery assumes a saturated population with constant $N$, and fitness involves a product of fertility and juvenile viability \citep[pp. 185]{crow_1970}. Unlike the Wright-Fisher model, generations can overlap in the lottery model. We generalize the lottery model to allow population density to vary, and show that this model can be interpreted as a density-dependent generalization of the Wright-Fisher model with overlapping generations.

@@

It is obvious from Eq.~\eqref{eq:LV} that selection can favor a superior competitor in a crowded population even if its saturation density is the same as, or lower than that of the other types present. However, note that the Lotka-Volterra model still closely couples selection to population density. Fig.~\ref{fig:LVvslottery}a shows Lotka-Volterra selection between two types with the same saturation density ($\alpha_{11}>\alpha_{22}$, $\alpha_{21}>\alpha_{12}$). Even though the initial and final densities of a sweep are the same, density is not constant during a sweep. Only a highly restricted subset of $r$ and $\alpha$ values will keep $N$ constant over a selective sweep (further details in Appendix C). Intuitively, for one type to exclude another with the same saturation density, inter-type competitive effects must be stronger than intra-type competitive effects, causing a dip in $N$ over the sweep.

@@

\section*{Type-specific responses to density and the strength of selection}

We are now in a position to analyze the validity of relative fitness models more explicitly. In the previous section we showed that selection and the regulation of population density can be completely independent of each other even if population growth is density-regulated, and moreover that MacArthur's argument for the $r$/$K$ scheme never precluded this possibility. Nevertheless, selection and density regulation \textit{are} intimately connected in widely used models of population growth. To understand why this poses a problem for Eq.~\eqref{eq:canonical}, consider the simple birth-death model \cite[pp. 20]{kostitzin_1939} 
\begin{equation}
\frac{d n_i}{dt}=(b_i -\delta_iN) n_i \label{eq:simplebirthdeath}
\end{equation}
where $\delta_i$ is the per-capita increase in type $i$'s mortality rate due to crowding (for simplicity, there are no deaths when uncrowded). Then, starting from a monomorphic population, the frequency of an adaptive $\delta$-variant $\delta_i\rightarrow \delta_i(1-\epsilon)$ obeys 
\begin{equation}
\frac{d p_i}{dt}=\epsilon \delta_i N p_i(1-p_i). \label{eq:Ndependentsweep}
\end{equation}
The selection coefficient $s=\epsilon \delta_i N$ thus depends on density (compare \cite[pp. 29]{crow_1970}). On the other hand, the frequency of an adaptive $b$-variant $b_i\rightarrow b_i(1+\epsilon)$ will exactly obey Eq.~\eqref{eq:canonical} with $s=\epsilon b_i$, independent of density.

@@

The reason that our lottery model behaves so differently from Eq.~\eqref{eq:simplebirthdeath} is that each type can respond differently to density via $c$ without affecting the regulation of population density. That is, our lottery model separates density-dependent selection from ``selection-dependent density'' \citep{prout_1980}, two distinct effects that are conflated in simple growth models with linearly-density-dependent per-capita growth rates. If types in the lottery model differ in $c$ as well as $b$ or $d$, then the lottery model will exhibit similar complications as Fig.~\ref{fig:strengthofselection}.

@@

%This  also seems to be why the density-dependent selection literature expended such considerable effort to showing that evolution under density-dependent selection optimizes population size in some sense \citep{roughgarden_1979}

It is important to recognize that MacArthur's argument does not imply a dichotomy between $r$-selection (uncrowded) and $K$-selection (crowded), with the latter taken to mean selection for greater saturation density. A better dichotomy is between interaction-dependent selection and interaction-independent selection. That is, selective shifts in frequency are a result of differences in absolute growth rates, but these differences can arise in two logically distinct ways: 1) some types expand more rapidly in the absence of interactions between individuals or 2) some types are superior in their interactions with other types. Population density is a key factor controlling whether individuals interact, thereby setting the relative contributions of these forms of selection.

%In the case of exploitation competition for consumable resources, intra- and inter-type competition are connected via resource use ``efficiency'', but obviously this need not be true more generally \citep{gill_1974,case_1974}.

%To avoid confusion with other forms of ''$K$-selection'', selection for competitive ability has been called ``$\alpha$-selection'' after the competition coefficients in the Lotka-Volterra equation \citep{gill_1974,case_1974,joshi_2001}. However, competitive success as measured by $\alpha$ (i.e. the per-capita effect of one type on another type's growth rate) is only partly determined by individual competitive ability --- in the presence of age-structured competition and territoriality, it also includes the ability of each type to produce contestants i.e. $b$ in our model.

One potential limitation of our model as a general-purpose model of density-dependent selection is its restriction to interference competition between juveniles for durable resources (lottery recruitment to adulthood), analogous to the ubiquitous assumption of viability selection in population genetics \citep[p. 45]{ewens_2004}. In some respects this is the complement of consumable resource competition models, which restrict their attention to indirect exploitation competition, typically without age structure \citep{tilman_1982}. In the particular case that consumable resources are spatially localized (e.g. due to restricted movement through soils), resource competition and territorial acquisition effectively coincide, and in principle resource competition could be represented by a competitive ability $c$ (or conversely, $c$ should be derivable from resource competition). The situation is more complicated if the resources are well-mixed, since, in general, resource levels then need to be explicitly tracked. It seems plausible that explicit resource tracking may not be necessary when the focus is on the evolution of similar types that use identical resources rather than the stable co-existence of widely differing species with different resource preferences \citep{ram_2016}. We are not aware of any attempts to delineate conditions under which explicit resource tracking is unnecessary even if it is assumed that community structure is ultimately determined by competition for consumable resources. More work is needed connecting resource competition models to the density-dependent selection literature, since most of the former has to date been focused on narrower issues of the role of competition at low resource availability and in the absence of direct interactions between organisms at the same trophic level \citep{aerts_1999,davis_1998,tilman_2007}.

While our model can be applied to species rather than types (e.g. ecological invasions), our focus is type evolution i.e. the change in allele frequencies over time. Our assumption that there are no large $c$ discrepancies (section ``Mean field approximation'') amounts to a restriction on the amount of genetic variation in $c$ in the population. Since beneficial mutation effect sizes will typically not be much larger than a few percent, large $c$ discrepancies can only arise if the mutation rate is extremely large, and so the assumption will not be violated in most cases. However, this restriction could become important when looking at species interactions rather than type evolution.

Another issue with the constant-$N$ relative fitness description of selection is that it precludes consideration of longer-term aspects of the interplay between evolution and ecology such as population extinction. Adaptive dynamics currently provides a powerful framework for adresing the complex feedbacks between evolutionary change and population density. However, the focus of adatpive dynamics is trait evolution rather than the underlying genetics, and in particular, selective sweeps are typically subsumed into effectively-instantaneous ``trait substitutions''. We emphasize that our focus here has been the description of selection itself, which is particularly critical for making sense of evolution at the genetic level.

%\section*{Acknowledgments}

%We thank Peter Chesson and Joachim Hermisson for many constructive comments on this manuscript. This work was financially supported by the National Science Foundation (DEB-1348262).

>>>>

\title{Density-dependent selection and the limits of relative fitness}
\author{Jason Bertram $^{1,\ast}$ \\ 
Joanna Masel $^{1}$}

@@

\textit{Keywords}: $r$/$K$ selection, eco-evolutionary synthesis.

@@

TBA

@@

In uncrowded populations, relative fitness simply represents differences in intrinsic growth rate. In discrete time, the change in frequency of type $i$ is $\Delta p_i=\left(\frac{W_i}{\overline{W}}-1\right) p_i$, where $W_i$ is the intrinsic absolute growth factor of type $i$,  $\overline{W}=\sum_i W_i p_i$ is the population mean $W$. Here we can rescale $W$ however we please and replace it with ``relative fitness'' $w$ without affecting the ratio $\frac{W_i}{\overline{W}}=\frac{w_i}{\overline{w}}$. In continuous time, the canonical selection equation is $\frac{d p_i}{dt}=(r_i-\overline{r}) p_i$, where $W$ is replaced by the intrinsic exponential growth rate $r$ \citep[pp. 26]{crow_1970}. If there are two types present, a wildtype $j$ and a mutant $i$ for instance, then the continuous time canonical selection equation can be written as
\begin{equation}
\frac{d p_i}{dt}=s p_i(1-p_i), \label{eq:canonical}
\end{equation}
where the constant selection coefficient $s$ is the difference in $r$ between types. The corresponding adaptive sweeps follow a logistic curve.

The difficulty with Eq.~\eqref{eq:canonical} arises in crowded populations. Since crowded and uncrowded conditions are so different, $s$ will often depend on density and cannot be treated as a simple constant (the possible frequency-dependence of $s$ is a separate issue, because this still leaves a closed model in terms of the frequencies). This issue is evaded by simply assuming that total population density $N$ has reached its equilibrium value, which is assumed to be a fixed constant. The selection coefficient $s$ now parameterizes the rate at which selection changes relative frequencies, but no longer corresponds to differences in intrinsic growth rates $r$.

Yet MacArthur famously showed that when population growth is density-regulated, selection in crowded populations is intimately connected to the ability to keep growing at higher densities than other types can tolerate \citep{macarthur_1967}. The classic example is the logistic model, where the type with the greatest saturation population density ``$K$'' excludes the others (Fig.~\ref{fig:Ksel}a). Similarly, the ``$R^*$ rule'', a central tenet of resource competition theory, states that when growth is limited by a single homogeneous consumable resource, the type able to deplete the resource to the lowest equilibrium density $R^*$ excludes the others \citep{grover_1997}. Differences in $R^*$ will often entail differences in saturation density. The Lotka-Volterra competition model also couples selection in crowded populations to changes in $N$ except in a few special cases (we return to this in section ``$K$-selection, $c$-selection and relative fitness''). In these examples, both $N$ and $s$ change during, and as a result of, adaptive sweeps. It would therefore seem that the ubiquitous constant-$N$, relative fitness description of selection is incompatible with a huge class of the population ecological processes driving selection (Fig.~\ref{fig:Ksel}b).

@@

The relative fitness description has been justified in broadly two different ways for crowded populations (we do not discuss Wagner's [\citeyear{wagner_2010}] measure-theoretical justification, which is explicitly independent of population biology and thus falls outside of our scope). The first is to assume that selection is independent of density but still allow density to be affected by selection \citep[pp. 468]{barton_2007} \citep{prout_1980}. While this  allows us to relax the assumption of constant $N$, it does not address the problem that $s$  can depend on density. In the examples from the previous paragraph, selection is density-dependent; indeed, the type-specific responses to density are at the center of MacArthur's argument and the density-dependent selection literature that grew out of it \citep{roughgarden_1979}.

Second, constant $N$ and $s$ can both be seen as a short-term linear approximation \cite[pp. 277]{ewens_2004}. That is, within a sufficiently short time frame, $N$ and $s$ can be treated as constant. Provided that selection is sufficiently weak and stable over time, this ``short-term''  assumption is not a major restriction. Yet it is increasingly recognized that selection is not always weak, that it can fluctuate considerably over time, and that $N$ can vary by orders of magnitude over a few generations as a routine feature of a population's ecology \citep{messer_2016}. These are not rare exceptions, but occur widely in nature and the lab, including in wild \textit{Drosophila} \citep{bergland_14}. Nevertheless, relative fitness models are the foundation for much of the population genetic literature, and are still widely used without considering the ``short-term'' restriction or the lack of integration with population ecology \citep{mallet_2012}. Thus, it is important to understand the population ecological basis of relative fitness models, both to gain insight into their domain of applicability, and as part of the broader challenge of synthesizing ecology and evolution.

Another issue with the constant-$N$ relative fitness description of selection is that it precludes consideration of longer-term aspects of the interplay between evolution and ecology such as population extinction. Adaptive dynamics currently provides a powerful framework for adresing the complex feedbacks between evolutionary change and population density \citep{diekmann_2004}. However, the focus of adaptive dynamics is trait evolution rather than the underlying genetics, and in particular, selective sweeps are typically subsumed into effectively-instantaneous ``trait substitutions''. We emphasize that our focus here is the description of selection itself, which is particularly critical for making sense of evolution at the genetic level.

Here we analyze the population ecology of relative fitness using a novel model of density-dependent population growth based on territorial contests. Rather attempting to make sense of relative fitness in the standard models of population growth mentioned above (e.g. \cite{mallet_2012}), we instead do the reverse, and attempt to make population ecological sense of the widely-used Wright-Fisher, constant-$N$, relative fitness model. Our starting point is the classic lottery model of territorial contest \citep{sale_77,chesson_1981}. Like the Wright-Fisher model, the classic lottery assumes a saturated population with constant $N$, and fitness involves a product of fertility and juvenile viability \citep[pp. 185]{crow_1970}. Unlike the Wright-Fisher model, generations can overlap in the lottery model. We generalize the lottery model to allow population density to vary, and show that this model can be interpreted as a density-dependent generalization of the Wright-Fisher model with overlapping generations.

@@

It is obvious from Eq.~\eqref{eq:LV} that selection can favor a superior competitor in a crowded population even if its saturation density is the same as, or lower than that of the other types present. However, note that the Lotka-Volterra model still closely couples selection to population density \citep{smouse_1976}. Fig.~\ref{fig:LVvslottery}a shows Lotka-Volterra selection between two types with the same saturation density ($\alpha_{11}>\alpha_{22}$, $\alpha_{21}>\alpha_{12}$). Even though the initial and final densities of a sweep are the same, density is not constant during a sweep. Only a highly restricted subset of $r$ and $\alpha$ values will keep $N$ constant over a selective sweep (further details in Appendix C). Intuitively, for one type to exclude another with the same saturation density, inter-type competitive effects must be stronger than intra-type competitive effects, causing a dip in $N$ over the sweep.

@@

\subsection*{Density-dependent selection and the strength of selection}

We are now in a position to analyze the validity of Eq.~\eqref{eq:canonical} more explicitly. In the previous section we showed that selection and the regulation of population density can be completely independent of each other even if population growth is density-regulated, and moreover that MacArthur's argument for the $r$/$K$ scheme never precluded this possibility. Nevertheless, selection and density regulation \textit{are} intimately connected in widely used models of population growth. To understand why this poses a problem for Eq.~\eqref{eq:canonical}, consider the simple birth-death model \cite[pp. 20]{kostitzin_1939} 
\begin{equation}
\frac{d n_i}{dt}=(b_i -\delta_iN) n_i \label{eq:simplebirthdeath}
\end{equation}
where $\delta_i$ is the per-capita increase in type $i$'s mortality rate due to crowding (for simplicity, there are no deaths when uncrowded). Then, starting from a monomorphic population, the frequency of an adaptive $\delta$-variant $\delta_i\rightarrow \delta_i(1-\epsilon)$ obeys 
\begin{equation}
\frac{d p_i}{dt}=\epsilon \delta_i N p_i(1-p_i). \label{eq:Ndependentsweep}
\end{equation}
The selection coefficient $s=\epsilon \delta_i N$ thus depends on density (compare \cite[pp. 29]{crow_1970}). On the other hand, the frequency of an adaptive $b$-variant $b_i\rightarrow b_i(1+\epsilon)$ will exactly obey Eq.~\eqref{eq:canonical} with $s=\epsilon b_i$, independent of density.

@@

@@

It is widely recognized that the evolutionary description of selection is poorly integrated with the ecological literature on population growth \citep{mallet_2012}. Typically this is cast as a narrow focus on frequencies on the part of evolutionary geneticists, who, perhaps out of necessity, omit the complexities of population ecology. While there may be some truth to this, we argue that part of the reason for the disconnect between the evolutionary geneticist's selection and the ecologist's selection is that relative fitness models actually contain important ecological processes omitted in many ecological models, including the classical Lotka-Volterra and logistic models. To illustrate, we highlight two important properties of our extended lottery model.

First, our lottery model contains a ``reproductive excess'' [any good suggestions for citations here, Joanna?] that allows selection (and competition) to occur among juveniles without affecting population density at the adult stage. This excess appears when the number of propagules produced is greater than the number of available territories. By contrast, in classical ecological growth models only one life-history stage is represented and selection is intrinsically connected to the effects of each type's adult density on the other. These are restrictive assumptions --- it is often the case that competition and selection are concentrated at the more fragile juvenile stage \citep{chesson_1983}. Reproductive excesses tend to produce strictly-relative lottery-type contests in which fitter types can grow at the expense of others by preferentially filling the available adult ``slots''. The number of slots can remain fixed or change independently of the large amount of selection occurring at the juvenile stage. Reproductive excesses expand the domain of validity of Eq.~\eqref{eq:canonical} by facilitating selection-independent density \citep{prout_1980}.

Second, density is regulated by fecundity $b$ and mortality $d$ traits that are affected equally by density. There are only so many territories, and all territories are equally available to all types. This removes the density-dependence that would make selection on $b$ and $d$ exhibit the difficulties shown in Fig.~\ref{fig:strengthofselection} without removing density-dependent selection from the model as whole via $c$ (by contrast, \cite[pp. 468]{barton_2007} allows density to depend on selection but removes density-dependent selection entirely). Although standard ecological models do have type-specific regulation of density (setting aside special parameter choices \citep{mallet_2012}), it is not obvious that we should expect this as a general rule. Much of our intuition about type-specific regulation of density (and density-dependent selection in general) is grounded in the logistic model and its diploid bi-allelic variants \citep{roughgarden_1979}, which should hardly be regarded as universal \citep{christiansen_2004,mallet_2012}. Consumable resource models are a better guide, because they explicitly model the processes underlying an important form of competition. Density regulation is type-specific in well-mixed resource competition models, but for many populations consumable resources are not well-mixed. Spatial localization of consumable resources (e.g. due to restricted movement of  nutrients through soils) will tend to create a territorial situation similar to the lottery model, where resource competition only occurs locally and is subsumed into the competitive ability $c$. Density regulation will then apply at the population level similarly to the lottery model, allowing $N$ to change via $b$ or $d$ without affecting Eq.~\eqref{eq:canonical}.

Relative fitness breaks down if strong selection occurs in a way that changes population density and is also density-dependent (Fig.~\ref{fig:strengthofselection}). In our lottery model, the latter can only occur if types differ in $c$ as well as $b$ or $d$. The reason for this is that density-dependent ($c$) and density-determining traits ($b$ and $d$) are completely separate.

It is interesting to compare this separation of traits with the $r$/$K$ scheme, often presented as a dichotomy between $r$-selection (uncrowded) and $K$-selection (crowded), with the latter taken to mean selection for greater saturation density (e.g. \citep{gill_1974}). As we have seen, MacArthur's argument does not justify this dichotomy at all. A better dichotomy is between interaction-dependent selection and interaction-independent selection. That is, selective shifts in frequency are a result of differences in absolute growth rates, but these differences can arise in two logically distinct ways: i) some types expand more rapidly in the absence of interactions between individuals or ii) some types are superior in their interactions with other types. In our the lottery model, $b$ and $d$ are interaction-independent traits (both within and between types), whereas $c$ is entirely about interactions between types. This logical distinction captures the essence of crowding, namely that crowding forces you to displace others in order to grow. Composite parameters like $K$ and $R^{*}$ conflate the regulation of density (factors which cause density trajectories to move ``outwards'' in Fig.~\ref{fig:Ksel}a) with interaction-dependent displacement (factors which cause density trajectories to move diagonally in Fig.~\ref{fig:Ksel}b).

Relative fitness models truly breaks down if $N$ is far from equilibrium and selection is density-dependent. For example, wild \textit{Drosophila} experience huge seasonal boom-bust cycles in population density coupled to strong selection that drives large swings in allele frequency \citep{bergland_14}. In this case there is no choice but to abandon relative fitness. Our density-dependent lottery model may not be a particularly good description of \textit{Drosophila} ecology, but the close connection between our model and Wright-Fisher is particularly useful, because drift in our model should behave broadly similarly. Thus, our model it should provide a useful starting point for analyzing evolution in this and other from-from-equilibrium situations.

\section*{Acknowledgments}

We thank Peter Chesson and Joachim Hermisson for many constructive comments on this manuscript. This work was financially supported by the National Science Foundation (DEB-1348262).
